---
title: "Speeding Up OSRM Isochrone"
author: "Ethan Roubenoff"
date: "10/3/2021"
output:
  md_document:
    variant: gfm
    preserve_yaml: yes
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding,
  output_file=paste0(Sys.Date(), "-", sub(".Rmd", ".md",inputFile)), output_dir =
  "~/eroubenoff.github.io/_posts") })
layout: post
tags:
- jekyll
- r-markdown
always_allow_html: yes

---
```{r, include=FALSE}
library(tidyverse)
base_dir <- "~/eroubenoff.github.io" # i.e. where the jekyll blog is on the hard drive.
base_url <- "/" # keep as is

# If the document is currently being knit, do this; skip it in normal execution
if (!is.null(knitr::current_input())){
  
  # Output path for figures
  fig_path <- paste0("assets/img/", str_remove(knitr::current_input(), ".Rmd"), "/")
  
  # Set base directories
  knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)
  
  # Set figure directories
  knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',
                      message=FALSE, warning=FALSE,
                      cache = FALSE)
}

```


For my work with [BIMI](https://bimi.berkeley.edu/mapping-spatial-inequality),
I am working on developing an index of service accessibility.
As part of this, I need to generate a lot of travel-time buffers around points:
basically, if you start at a location and drive the speed limit in any direction,
what is the polygon that you can access? These are called isochrones,
and I'm using the [Open Source Routing Machine](http://project-osrm.org/) and [OpenStreetMap](http://project-osrm.org/) data
to generate them.

Eventually, when the research is a little further along, I'll do a post on 
how this process works---I learned a lot setting up OSRM and OSM and it's 
worth putting it out on the internet for any OSRM newbies.

This post is specifically about how I sped up the OSRM function `osrmIsochrone`.
The package function is optimized for large-area accuracy (I think), but since
our buffers are relatively small (30 minutes driving) we can let go of
one specific thing.  No shade on the OSRM developers, but I was able to speed
this up quite a lot.

Our setup: We are putting a 30-minute driving buffer around 100
points randomly dropped in California:

```{r, }
library(sf)
library(tmap)
set.seed(4949)
CA_shp <- tigris::states(cb = TRUE, class = "sf") %>%
  filter(NAME == "California")
CA_pts <- st_set_geometry(data.frame(id = 1:100), st_sample(CA_shp, 100))
tm_shape(CA_shp) +
  tm_polygons() +
  tm_shape(CA_pts) + 
  tm_dots()
```


Let's set up OSRM:
```{r}
library(osrm)
options(osrm.server = "http://0.0.0.0:5000/", osrm.profile = "driving")
```


I find it easiest for parallelization (which I'll get to later)
to write the process of generating isochrones in a function:

```{r}
fx_base <- function(i, pts) {
  message(paste("Running location", i))
  tryCatch({
    return(st_set_geometry(pts[i,],
                           osrm::osrmIsochrone(
                             loc = pts[i,],
                             breaks = c(30),
                             res = 60,
                             returnclass =  "sf") %>% 
                             filter(max == 30) %>% 
                             st_geometry()))
  }, error = function(e) {
    message(paste("Failure at location", i, ":", e))
    print(pts[i, ])
    return()
  })
}

```

What this function does is take an index (as you'll see, we'll loop
through `1:nrow(CA_pts)`) and the `CA_pts` object, get the isochrone
around that point, and return the original entry with the new geometry.
The `osrmIsochrone` function doesn't preserve any of the attributes
passed, so this method returns the attributes of the original 
point with the new buffer.  We will eventually `rbind` them all together.


```{r, message=TRUE}
library(tictoc)
tic()
suppressMessages({
  CA_isochrones_base <- map(1:nrow(CA_pts), ~fx_base(., CA_pts))
  CA_isochrones_base <- data.table::rbindlist(CA_isochrones_base) %>% st_as_sf
  })

isochrones_base_time <- toc()
```


This took about 95 seconds to run 100 points.  This is fine for a small
operation, but really a pain for a bigger one! 
Some values will fail. That's okay-- debugging that is a pain and ususally just
means that there's no road that gets to that point. And I know it's weird
that I'm using tidyverse syntax on SF with a data.table command-- it works, don't fight it!

Here's what they look like:

```{r}
tm_shape(CA_shp) +
  tm_polygons() +
  tm_shape(CA_isochrones_base) + 
  tm_polygons(col = "green", alpha= 0.7)
```


I can't figure out how to put code profiling in an r markdown document (at least)
one that is static, but the output of `profvis` makes it look like `st_make_grid`
is eating up a lot of computation time.  Some things take longer, like i/o and
`fromJSON`, but I can't really do anything about those.

My understanding of the algorithm is that it takes the input point, creates
a grid of points on top, and evaluates the time from the input point
to each point; it then draws a contour around all points that are below
a certain travel time. However, the way it does that grid of points is markedly
ineffecient. From the source code (many things are cut out):

```{r, eval = FALSE}
function (loc, breaks = seq(from = 0, to = 60, length.out = 7), 
    exclude = NULL, res = 30, returnclass = "sp") {
  
  ...
  ...
  
    if (options("osrm.profile") == "walk") {
        speed = 10 * 1000/60
    }
    if (options("osrm.profile") == "bike") {
        speed = 20 * 1000/60
    }
    if (options("osrm.profile") == "driving") {
        speed = 130 * 1000/60
    }
    dmax <- tmax * speed
    sgrid <- rgrid(loc = loc, dmax = dmax, res = res)
    lsgr <- nrow(sgrid)
    f500 <- lsgr%/%300
    r500 <- lsgr%%300
    listDur <- list()
    listDest <- list()
    if (getOption("osrm.server") != "http://router.project-osrm.org/") {
        sleeptime <- 0
    }
    else {
        sleeptime <- 1
    }
    if (f500 > 0) {
        for (i in 1:f500) {
            st <- (i - 1) * 300 + 1
            en <- i * 300
            dmat <- osrmTable(src = loc, dst = sgrid[st:en, ], 
                exclude = exclude)
            durations <- dmat$durations
            listDur[[i]] <- dmat$durations
            listDest[[i]] <- dmat$destinations
            Sys.sleep(sleeptime)
        }
        if (r500 > 0) {
            dmat <- osrmTable(src = loc, dst = sgrid[(en + 1):(en + 
                r500), ], exclude = exclude)
            listDur[[i + 1]] <- dmat$durations
            listDest[[i + 1]] <- dmat$destinations
        }
    }
    else {
        dmat <- osrmTable(src = loc, dst = sgrid, exclude = exclude)
        listDur[[1]] <- dmat$durations
        listDest[[1]] <- dmat$destinations
    }
    ...
    ...
    
}
```



The long and short of what I did is cut out this loop that contains `for (i in 1:f500) ...`:
```{r, eval =FALSE}

er_osrmIsochrone <- function(loc, breaks = c(30), exclude = NULL, res = 30, returnclass = "sf"){
  ...
  ...
  
   rgrid <- function(loc, dmax, res){
    # create a regular grid centerd on loc
    boxCoordX <- seq(from = sf::st_coordinates(loc)[1,1] - dmax,
                     to = sf::st_coordinates(loc)[1,1] + dmax,
                     length.out = res)
    boxCoordY <- seq(from = sf::st_coordinates(loc)[1,2] - dmax,
                     to = sf::st_coordinates(loc)[1,2] + dmax,
                     length.out = res)
    sgrid <- expand.grid(boxCoordX, boxCoordY)
    sgrid <- data.frame(ID = seq(1, nrow(sgrid), 1),
                        COORDX = sgrid[, 1],
                        COORDY = sgrid[, 2])
    sgrid <- sf::st_as_sf(sgrid,  coords = c("COORDX", "COORDY"),
                          crs = st_crs(loc), remove = FALSE)
    return(sgrid)
  }
  
  oprj <- st_crs(loc)
  loc <- loc[1,]
  loc <- st_transform(loc, 3857)
  row.names(loc) <- "0"
  
  # max distance mngmnt to see how far to extend the grid to get measures
  breaks <- unique(sort(breaks))
  tmax <- max(breaks)
  if(options('osrm.profile')=="walk"){
    speed =  10 * 1000/60
  }
  if(options('osrm.profile')=="bike"){
    speed =  20 * 1000/60
  }
  if(options('osrm.profile')=="driving"){
    speed =  130 * 1000/60
  }
  dmax <- tmax * speed
  sgrid <- rgrid(loc = loc, dmax = dmax, res = res)
  
  # dmat <- osrmTable(src = loc, dst = sgrid, exclude = exclude)
  dmat <- osrmTable(src = loc, dst = sgrid)
  durations <- dmat$durations
  destinations <- dmat$destinations
  ...
  ...
}
```

There are a few other operations, and you can view the full source code [here]("assets/r_markdown/er_osrmIsochrone.R").

Let's test how fast it is:
```{r}
source("er_osrmIsochrone.R")
fx_er <- function(i, pts) {
  message(paste("Running location", i))
  tryCatch({
    return(st_set_geometry(pts[i,],
                           er_osrmIsochrone(
                             loc = pts[i,],
                             breaks = c(30),
                             res = 60,
                             returnclass =  "sf") %>% 
                             filter(max == 30) %>% 
                             st_geometry()))
  }, error = function(e) {
    message(paste("Failure at location", i, ":", e))
    print(pts[i, ])
    return()
  })
}

tic()
suppressMessages({
  CA_isochrones_base <- map(1:nrow(CA_pts), ~fx_er(., CA_pts))
  CA_isochrones_base <- data.table::rbindlist(CA_isochrones_base) %>% st_as_sf
  })

isochrones_er_time <- toc()
```


That's not a bad speed up! 74 seconds; saved a few. However, the 
real advantage comes when you have thousands of points and want to run them
in parallel:



```{r}
library(parallel)
CA_pts <- st_set_geometry(data.frame(id = 1:10000), st_sample(CA_shp, 10000))
tic()
  out <- suppressMessages({mclapply(1:10000, function(i) {fx_base(i, CA_pts)}, mc.cores = 16)})
isochrones_base_parallel_time <- toc()

tic()
  out <- suppressMessages({mclapply(1:10000, function(i) {fx_er(i, CA_pts)}, mc.cores = 16)})
isochrones_er_parallel_time <- toc()

```



In this toy case it's not a huge speed up, but on our applied problem,
I was able to cut a 2-hour method down to 30 mintes. And that's not nothing!



